Eco-evolutionary dynamics of finite populations from demographic first principles

### 2023-01-31 ind_files/03_BD_processes.tex
Add section on stochastic trait freq dynamics

\myfig{0.9}{Media/3.2_BD_process_2D.png}{\textbf{Schematic description of an $m$-dimensional birth-death process. (A)} Consider a population of birds in which individuals are either red or blue. In this case, we have $m=2$, since there are two types of individuals in the population. \textbf{(B)} The state of the system can be described by a vector containing the number of individuals of each discrete type (in this case, the number of red and blue birds in the population). Births and deaths result in changes in the elements of the state vector.}{fig_nD_pop_description}

@@

\subsection{The system-size expansion}
As before, we now assume we can find a system size parameter $K > 0$ such that we can make the substitutions
\begin{align*}
    \mathbf{x} &= \frac{\mathbf{v}}{K}\\
    b^{(K)}_i(\mathbf{x}) &= \frac{1}{K}b_i(\mathbf{v})\\
    d^{(K)}_i(\mathbf{x}) &= \frac{1}{K}d_i(\mathbf{v})
\end{align*}
and define new step operators $\Delta_{i}^{\pm}$ by their action on any real-valued function $f(\mathbf{x},t)$ as
\begin{equation}
\label{nD_step_operators_rescaled}
    \Delta_{i}^{\pm}f([x_1,\ldots,x_m],t) = f([x_1,\ldots,x_i \pm \frac{1}{K}, \ldots x_m],t)
\end{equation}
In terms of these new variables, \eqref{nD_M_eqn} becomes
\begin{equation}
\label{nd_M_eqn_rescaled}
\frac{\partial P}{\partial t}(\mathbf{x},t) = K\sum\limits_{j=1}^{m}\left[(\Delta_j^{-}-1)b^{(K)}_j(\mathbf{x})P(\mathbf{x},t) + (\Delta_j^{+}-1)d^{(K)}_j(\mathbf{x})P(\mathbf{x},t)\right]
\end{equation}
If $K$ is large, we can once again Taylor expand the action of the step operators as
\begin{equation*}
f([x_1,\ldots,x_i \pm \frac{1}{K}, \ldots x_m],t) = f(\mathbf{x},t) \pm \frac{1}{K}\frac{\partial f}{\partial x_i}(\mathbf{x},t) + \frac{1}{2K^2}\frac{\partial^2f}{\partial x_i^2}(\mathbf{x},t) + \mathcal{O}(K^{-3})
\end{equation*}
which, after substituting into \eqref{nd_M_eqn_rescaled}, yields the equation
\begin{equation}
\label{nD_FPE}
\setlength{\fboxsep}{2\fboxsep}\boxed{\frac{\partial P}{\partial t}(\mathbf{x},t) = \sum\limits_{j=1}^{m}\left[-\frac{\partial}{\partial x_j}\{A_j^{-}(\mathbf{x})P(\mathbf{x},t)\} + \frac{1}{2K}\frac{\partial^2}{\partial x_j^2}\{A_j^{+}(\mathbf{x})P(\mathbf{x},t)\}\right]}
\end{equation}
where
\begin{equation*}
A_{i}^{\pm}(\mathbf{x}) = b^{(K)}_i(\mathbf{x})\pm d^{(K)}_i(\mathbf{x})
\end{equation*}
Equation \eqref{nD_FPE} is an $m$-dimensional Fokker-Planck equation, and corresponds to the $m$-dimensional It\^o process
\begin{equation}
\label{nD_Ito_SDE}
    d\mathbf{X}_{t} = \mathbf{A^-}(\mathbf{X}_t)dt + \frac{1}{\sqrt{K}}\mathbf{D}(\mathbf{X}_t)d\mathbf{B}_t
\end{equation}
where $\mathbf{A^-}(\mathbf{X}_t)$ is the $m$ dimensional `drift vector' with $i$\textsuperscript{th} element $ = A^{-}_{i}(\mathbf{X}_t)$. $\mathbf{D}(\mathbf{X}_t)$ is the $m \times m$ `diffusion matrix' and satisfies $\mathbf{D}\mathbf{D}^{\mathrm{T}} = \mathbf{A^+}(\mathbf{X}_t)$, where $\mathbf{A^+}(\mathbf{X}_t))$ is an $m$ dimensional vector with $i$\textsuperscript{th} element $ = A^{+}_{i}(\mathbf{X}_t)$. Finally, $\mathbf{B}_t$ is the $m$-dimensional Brownian motion and can be thought of as a vector of independent one-dimensional Brownian motions (which have been defined in \ref{intro_SDE}). This is the `mesoscopic' description of our process.

@@

In this multi-dimensional case, the macroscopic description amounts to a description of evolutionary game theory under mild assumptions on the birth and death rate vectors. We show this below, adopting some methods first outlined by \citep{page_unifying_2002}. Let us assume that we can write the birth and death rate vectors as

@@

Where we have used the notation $Q_i(\mathbf{p}) = Q_i(\mathbf{x})/N_K(t)$ for notational clarity. The first term of \eqref{nD_replicator_mutator} describes changes due to faithful (non-mutational) replication, and the second describes changes due to mutation. For this reason, equation \eqref{nD_replicator_mutator} is called the \emph{replicator-mutator equation} in the evolutionary game theory literature, where the individual `types' are interpreted to be pure strategies. If in addition, each $w_i(\mathbf{x})$ is linear in $\mathbf{x}$, meaning we can write $w_i(\mathbf{x}) = \sum_{j}a_{ij}x_j$ for some set of constants $a_{ij}$, then we get the replicator-mutator equation for matrix games, and the constants $a_{ij}$ form the `payoff matrix'. As is well-known, the replicator equation (without mutation) for matrix games with $m$ pure strategies is equivalent to the generalized Lotka-Volterra equations for a community with $m-1$ species \citep{hofbauer_evolutionary_1998}, providing the connection to community ecology.  Equation \eqref{nD_replicator_mutator} is also equivalent to Eigen's \emph{quasispecies equation} from molecular evolution \citep{page_unifying_2002} if each `type' is interpreted as a genetic sequence and each $w_i(\mathbf{x})$ is a constant function\footnote{Mutational effects are often additionally assumed to act through direct `transmission probabilities' of mutating from one type to another. This means that we can write $Q_i(\mathbf{p}) = \sum\limits_{j\neq i}Q_{ij}p_j$, where $Q_{ij} \geq0$ is a constant describing the probability of a $j \to i$ mutation (conditioned on the occurrence of a mutation). Substituting this into \eqref{nD_replicator_mutator} yields an equation in terms of `$Q$-matrices' or `mutation matrices' that may be more familiar to some biologists.}. We can now calculate how the mean of any `trait level' quantity $f$, defined as $f_i$ for the $i$\textsuperscript{th} trait, changes in the population (For example, setting $f_i = x_i$ gives us the mean trait value in the population). Multiplying both sides of equation \eqref{nD_replicator_mutator} by $f_i$ and summing over all $i$, we obtain

@@

@@

Let the deterministic trajectory obtained by solving \eqref{nD_det_limit} be given by $\mathbf{a}(t)$.  We can once again track 

@@

where $\delta_{ij}$ is the Kronecker delta symbol, defined by
\delta_{ij} = 
1 & i=j\\
0 & i\neq j

@@

&= y_iw_i(\mathbf{a}) + a_iD_{\mathbf{y}}(w_i(\mathbf{a})) + \mu D_{\mathbf{y}}(Q_i(\mathbf{a}))

@@

\label{nD_WNA_directional_derivative_for_replicator_eqns}

>>>>

\myfig{0.9}{Media/3.2_BD_process_2D.png}{\textbf{(A)} Consider a population of birds in which individuals are either red or blue. In this case, we have $m=2$, since there are two types of individuals in the population. \textbf{(B)} The state of the system can be described by a vector containing the number of individuals of each discrete type (in this case, the number of red and blue birds in the population). Births and deaths result in changes in the elements of the state vector.}{fig_nD_pop_description}

@@

\subsection{The system-size expansion}
As before, we now assume we can find a system size parameter $K > 0$ such that we can make the substitutions
\begin{align*}
\mathbf{x} &= \frac{\mathbf{v}}{K}\\
b^{(K)}_i(\mathbf{x}) &= \frac{1}{K}b_i(\mathbf{v})\\
d^{(K)}_i(\mathbf{x}) &= \frac{1}{K}d_i(\mathbf{v})
\end{align*}
and define new step operators $\Delta_{i}^{\pm}$ by their action on any real-valued function $f(\mathbf{x},t)$ as
\begin{equation}
\label{nD_step_operators_rescaled}
\Delta_{i}^{\pm}f([x_1,\ldots,x_m],t) = f([x_1,\ldots,x_i \pm \frac{1}{K}, \ldots x_m],t)
\end{equation}
In terms of these new variables, \eqref{nD_M_eqn} becomes
\begin{equation}
\label{nd_M_eqn_rescaled}
\frac{\partial P}{\partial t}(\mathbf{x},t) = K\sum\limits_{j=1}^{m}\left[(\Delta_j^{-}-1)b^{(K)}_j(\mathbf{x})P(\mathbf{x},t) + (\Delta_j^{+}-1)d^{(K)}_j(\mathbf{x})P(\mathbf{x},t)\right]
\end{equation}
If $K$ is large, we can once again Taylor expand the action of the step operators as
\begin{equation*}
f([x_1,\ldots,x_i \pm \frac{1}{K}, \ldots x_m],t) = f(\mathbf{x},t) \pm \frac{1}{K}\frac{\partial f}{\partial x_i}(\mathbf{x},t) + \frac{1}{2K^2}\frac{\partial^2f}{\partial x_i^2}(\mathbf{x},t) + \mathcal{O}(K^{-3})
\end{equation*}
which, after substituting into \eqref{nd_M_eqn_rescaled}, yields the equation
\begin{equation}
\label{nD_FPE}
\setlength{\fboxsep}{2\fboxsep}\boxed{\frac{\partial P}{\partial t}(\mathbf{x},t) = \sum\limits_{j=1}^{m}\left[-\frac{\partial}{\partial x_j}\{A_j^{-}(\mathbf{x})P(\mathbf{x},t)\} + \frac{1}{2K}\frac{\partial^2}{\partial x_j^2}\{A_j^{+}(\mathbf{x})P(\mathbf{x},t)\}\right]}
\end{equation}
where
\begin{equation*}
A_{i}^{\pm}(\mathbf{x}) = b^{(K)}_i(\mathbf{x})\pm d^{(K)}_i(\mathbf{x})
\end{equation*}
Equation \eqref{nD_FPE} is an $m$-dimensional Fokker-Planck equation, and corresponds to the $m$-dimensional It\^o process
\begin{equation}
\label{nD_Ito_SDE}
d\mathbf{X}_{t} = \mathbf{A^-}(\mathbf{X}_t)dt + \frac{1}{\sqrt{K}}\mathbf{D}(\mathbf{X}_t)d\mathbf{B}_t
\end{equation}
where $\mathbf{A^-}(\mathbf{X}_t)$ is the $m$ dimensional `drift vector' with $i$\textsuperscript{th} element $ = A^{-}_{i}(\mathbf{X}_t)$. $\mathbf{D}(\mathbf{X}_t)$ is the $m \times m$ `diffusion matrix' with $ij$th element $\left(\mathbf{D}(\mathbf{X}_t)\right)_{ij} = \delta_{ij}\left(A^{+}_{i}A^{+}_{j}\right)^{\frac{1}{4}}$, where $\delta_{ij}$ is the Kronecker delta symbol, defined by
\begin{equation*}
\delta_{ij} = 
\begin{cases}
1 & i=j\\
0 & i\neq j
\end{cases}
\end{equation*}
Finally, $\mathbf{B}_t$ is the $m$-dimensional Brownian motion and can be thought of as a vector of independent one-dimensional Brownian motions (which have been defined in \ref{intro_SDE}). This is the `mesoscopic' description of our process.

@@

In this multi-dimensional case, the macroscopic description amounts to a description of evolutionary game theory under mild assumptions on the birth and death rate vectors. We show this below, using methods first outlined by \citep{page_unifying_2002}. Let us assume that we can write the birth and death rate vectors as

@@

Where we have used the notation $Q_i(\mathbf{p}) = Q_i(\mathbf{x})/N_K(t)$ for notational clarity. The first term of \eqref{nD_replicator_mutator} describes changes due to faithful (non-mutational) replication, and the second describes changes due to mutation. For this reason, equation \eqref{nD_replicator_mutator} is called the \emph{replicator-mutator equation} in the evolutionary game theory literature, where the individual `types' are interpreted to be pure strategies. If in addition, each $w_i(\mathbf{x})$ is linear in $\mathbf{x}$, meaning we can write $w_i(\mathbf{x}) = \sum_{j}a_{ij}x_j$ for some set of constants $a_{ij}$, then we get the replicator-mutator equation for matrix games, and the constants $a_{ij}$ form the `payoff matrix'. As is well-known, the replicator equation (without mutation) for matrix games with $m$ pure strategies is equivalent to the generalized Lotka-Volterra equations for a community with $m-1$ species\citep{hofbauer_evolutionary_1998}, providing the connection to community ecology.  Equation \eqref{nD_replicator_mutator} is also equivalent to Eigen's \emph{quasispecies equation} from molecular evolution if each `type' is interpreted as a genetic sequence and each $w_i(\mathbf{x})$ is a constant function\footnote{Mutational effects are often additionally assumed to act through direct `transmission probabilities' of mutating from one type to another. This means that we can write $Q_i(\mathbf{p}) = \sum\limits_{j\neq i}Q_{ij}p_j$, where $Q_{ij} \geq0$ is a constant describing the probability of a $j \to i$ mutation (conditioned on the occurrence of a mutation). Substituting this into \eqref{nD_replicator_mutator} yields an equation in terms of `$Q$-matrices' or `mutation matrices' that may be more familiar to some biologists.}. We can now calculate how the mean of any `trait level' quantity $f$, defined as $f_i$ for the $i$\textsuperscript{th} trait, changes in the population (For example, setting $f_i = x_i$ gives us the mean trait value in the population). Multiplying both sides of equation \eqref{nD_replicator_mutator} by $f_i$ and summing over all $i$, we obtain

@@

\subsection{Stochastic trait frequency dynamics for finite populations}

Following ideas similar in spirit to \citep{mcleod_social_2019}, we can also use \eqref{nD_Ito_SDE} to write down stochastic equations for $p_i(t)$, the frequency of type $i$ individuals in the population to contrast with the replicator-mutator equation \eqref{nD_replicator_mutator} derived above. In appendix \ref{App_density_to_freq}, we show that we can use It\^{o}'s lemma to write  down a general stochastic equation for the frequencies of each type in the population.  Unlike \cite{mcleod_social_2019}, we make no assumptions about separations of time scales or strength of selection and are able to present an entirely general calculation. Letting $\tau_i(\mathbf{x}) = b^{(\textrm{int})}_{i}(\mathbf{x}) + d^{(\textrm{int})}_i(\mathbf{x})$ be the \emph{per-capita turnover rate} of the $i\textsuperscript{th}$ type and denoting $\overline{\tau} = \sum_i p_i\tau_i$, we show in appendix \ref{App_density_to_freq} that when the birth and death rates are defined by \eqref{nD_functional_forms_for_replicator}, then the frequency of the $i\textsuperscript{th}$ type in the population $\mathbf{X}_t$, changes according to the equation:
\label{nD_eqn_for_frequencies}
dp_i(t) &= \underbrace{\left[(w_i(\mathbf{x}) - \overline{w})p_i + \mu\left\{Q_i(\mathbf{p}) - p_i\left(\sum\limits_{j=1}^{m}Q_j(\mathbf{p})\right)\right\}\right]dt}_{\substack{\text{Infinite population predictions} \\ \text{(Replicator-mutator, Price, etc.)}}}\\
&- \frac{1}{K}\underbrace{\frac{1}{N(t)}\left[(\tau_i(\mathbf{x}) - \overline{\tau})p_i + \mu\left\{Q_i(\mathbf{p}) - p_i\left(\sum\limits_{j=1}^{m}Q_j(\mathbf{p})\right)\right\}\right]dt}_{\substack{\text{Directional finite population effects}\\\text{due to differential turnover rates}}}\\
&+ \frac{1}{\sqrt{K}}\underbrace{\left[\left(A^{+}_{i}\right)^{1/2}dB^{(i)}_t - p_i\sum\limits_{j=1}^{m}\left(A^{+}_{j}\right)^{1/2}dB^{(j)}_t\right]}_{\substack{\text{Non-directional finite population effects}\\\text{due to stochastic fluctuations}}}
where $B^{(1)}_t,B^{(2)}_t, \ldots, B^{(m)}_t$ are $m$ independent one-dimensional standard Brownian motion processes. The first term in this expression is the same as equation \eqref{nD_replicator_mutator} and describes directional changes in the population composition due to `classical' evolutionary forces such as selection and mutation. As we can see, the replicator-mutator equation is generically satisfied for all birth-death processes that are in the (very general) functional forms \eqref{nD_functional_forms_for_replicator}. The second term is an additional directional force on population composition that acts exactly analogously to the replicator equation but in the opposite direction (A higher relative $\tau_i$ leads to a lower frequency), and furthermore, is particular to finite populations. This term describes a biasing effect due to differential turnover rates and can intuitively be understood in the following sense: If a type $i$ has a higher $\tau_i$, it experiences greater turnover due to a generally higher birth and death rate and thus greater demographic stochasticity. Types with a higher $\tau_i$ thus tend to be eliminated by a stochastic analog of selection because they experience more chance events (births and deaths) in a given time period (notice the minus sign before the second term in \eqref{nD_eqn_for_frequencies}). This effect is less visible if the total population size is higher because larger populations generally experience less stochasticity, and this is reflected in the $1/N$ factor in this term. This stochastic analog of selection for reduced turnover rates is the force responsible for the `reversal of the direction of deterministic selection' observed in previous studies \citep{constable_demographic_2016, mcleod_social_2019} and is an effect that is only seen in finite populations. Note that types that tend to increase the \emph{total} population size (such as altruists in evolutionary theory and mutualists in ecological communities) experience a smaller negative effect due to differential turnover rates than types that do not facilitate such an increase (such as cheaters and highly competitive species), which could explain why mutualists and cooperators routinely persist for long periods in finite population models despite extinction being predicted in infinite population models. The fact that the entire term is multiplied by $(KN(t))^{-1}$, however, suggests that the effect of this force is weak for medium to large populations, which explains why the aforementioned persistence of cooperators is often only observed in restrictive sounding conditions such as quasi-neutrality, rapid attraction to a slow manifold, or a weak selection limit. In all three of these cases, the first term on the RHS of \eqref{nD_eqn_for_frequencies} becomes identically 0 and therefore no longer contributes to the total dynamics, thus revealing the effects of differential turnover rate.\\
Finally, the last term describes non-directional stochastic effects due to fluctuations and has a `spreading effect'. To the highest order in $\sqrt{K}$, the effect of these fluctuations can be described through the weak noise approximation, as we explain below. However, the $1/K$ factor in the second term of \eqref{nD_eqn_for_frequencies} means that the effect of differential turnover rates disappears in this weak noise approximation  (and also in other such approximate treatments, such as the Chemical langevin).

@@

If the noise is \emph{weak}, we can go a little further, as in the one-dimensional case. Let the deterministic trajectory obtained by solving \eqref{nD_det_limit} be given by $\mathbf{a}(t)$.  We can once again track 

@@

@@

&= y_iw_i(\mathbf{a}) + a_iD_{\mathbf{y}}(w_i(\mathbf{a})) + \mu D_{\mathbf{y}}(Q_i(\mathbf{a}))\label{nD_WNA_directional_derivative_for_replicator_eqns}

@@

\label{nD_moment_eqn_mean_replicator}
